\scuba: iterative map-making with the Sub-Millimetre User Reduction Facility

### 2013-01-07 smurf_paper.tex
screwed these up in 3f83b0cb0554

\caption{A comparison between single bolometer time-series in each of
  the 450 and 850\,\micron\ bands with the mixing chamber temperature,
  the line-of-sight 450\,\micron\ opacity derived from the JCMT water
  vapour monitor, and azimuth/elevation pointing offsets, before and
  after upgrading the instrument. The grey signals over-plotted in the
  top two panels show the residual time-series after removing the
  common-mode signals.  (a) Data taken during the S2SRO period,
  observation 29 on 2010-03-13. There is a strong correlation between
  the bolometers and the roughly $\sim$25\,s oscillation in the
  fridge, but only a minor correlation with the opacity and telescope
  motion. The nearly flat common-mode subtracted signals show: (i)
  that most of the low-frequency signal is common to all of the
  bolometers; and (ii) the non-correlated, and predominantly white
  noise at 450\,\micron\ is significantly larger than at
  850\,\micron. (b) Data taken with the fully-commissioned instrument,
  observation 38 on 2011-11-12. Unlike the S2SRO data, there is no
  strong signal produced by variations in the fridge temperature. The
  450\,\micron\ data have minimal correlated low-frequency noise (as
  evidenced by the lack of a difference between the raw and
  common-mode subtracted signals), although the 850\,\micron\ data
  show a common signal correlated with the opacity and telescope
  motion (note that this scan has a significantly larger amplitude
  than the S2SRO data set), mostly likely caused by a combination of
  changing airmass (anti-correlated with opacity), and magnetic field
  pickup (see Section~\ref{sec:magpickup} and
  Fig.~\ref{fig:magpickup}).}
\label{fig:bolos_mix}
\end{figure*}

@@

In the S2SRO data (Fig.~\ref{fig:bolos_mix}a), observation 29 on
2010-03-13, both bolometers share significant long-timescale structure
($\gsim10$\,s) that appears to be related to variations in the fridge
base temperature, although the similarity is clearly greater at
850\,\micron. Note that the total power in the fluctuations at
450\,\micron\ are comparable to those at 850\,\micron\ as one might
expect if there is a comparable varying thermal load from the fridge
at each wavelength that dominates. We also note that there is no
obvious strong correlation between the low-frequency signal structure,
at either wavelength, with the telescope motion. However, there is a
suggestion that the shorter-timescale behaviour of the opacity is
anti-correlated with the elevation, as expected.

@@

Data from the fully commissioned instrument, observation 38 on
2011-11-12, are shown in Fig.~\ref{fig:bolos_mix}b. Having solved the
fridge oscillation problem, these data no longer exhibit a correlation
with the mixing chamber (however, note that variations are still seen
in the mixing chamber signal; such variations do not necessarily
reflect changes in the focal plane temperature). There is minimal
correlated low-frequency noise in these 450\,\micron\ data, resulting
in little difference between the raw and common-mode subtracted
data. The 850\,\micron\ channel, however, exhibits a more significant
signal that is obviously correlated with the telescope motion. It is
also correlated amongst many of the detectors, and common-mode removal
corrects it to a large extent. Part of the reason that this is seen
here, and not in the S2SRO data set shown, is that the amplitude of
the scan pattern is larger. This ``scan-synchronous'' noise is
attributed to a combination of magnetic field pickup, as described in
Section~\ref{sec:magpickup}, and sky brightness variations due to
changes in elevation (not to be confused with underlying changes in
the atmosphere; a homogeneous atmosphere will appear brighter with
increasing airmass). Also similar to the S2SRO data, an
anti-correlation between the opacity and elevation is apparent.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{magpickup}
\caption{Evidence for significant magnetic field pickup for
  observation 16 on 2010-02-28 at 450\,\micron\ (s4a subarray).  The top
  panel shows two un-flatfielded (but mean-subtracted and step
  corrected) bolometer time-series from the same column, with a 200
  sample boxcar smooth (approximately 1\,s), illustrating that they
  are dominated by a similar signal with opposite signs. The second
  panel shows the dark squid signal for the column, also
  mean-subtracted and with the same boxcar smooth. The bottom panel
  shows the azimuthal and elevation offsets from the map centre (mean
  azimuth and elevation 171.9$^\circ$ and 68.0$^\circ$,
  respectively). Only the azimuthal signal is obviously correlated
  with the dark squids and bolometer signals, which suggests a
  magnetic field stationary with respect to the telescope dome as the
  source, since its direction with respect to the cryostat only
  changes with azimuthal motion.}
\label{fig:magpickup}
\end{figure}

@@

Example data from the 450\,\micron\ subarray s4a where pickup appears
to be significant (observation 16 on 2010-02-28) are shown in
Fig.~\ref{fig:magpickup}. The time-series for two bolometers in the
same column (not flatfielded) show that there is a strong signal with
a similar shape, but opposite signs. This behaviour is seen across the
entire array. The dark squid signal for the same columns exhibits a
similar shape and amplitude. Since the dark squid has no thermal
absorber or TES attached to it, this observed signal is not likely to
be optical or thermal in nature (although there can be some cross-talk
with the bolometers). Due to the fact that the sign of the gain in
each stage of SQUID amplification is random (although the combined
gain is constrained to be negative), and since magnetic field pickup
is only seen at the input to the second stage, the pickup can appear
with random signs for bolometers along a column, giving it a distinct
signature from other common signals that always appear with the same
sign.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{skynoise}
\caption{Illustration of the dominance of atmospheric variations on
  long-timescales using 35\,min. of data taken in highly-variable
  weather conditions (line-of-sight 450\,\micron\ opacity ranging from
  3.6 to 7.7) from the fully-commissioned instrument, observation 18
  on 2011-03-24. Red and blue lines show the 850 and 450\,\micron\
  common-mode signals, corresponding to the left- and right- vertical
  axes, respectively. The black line shows the line-of-sight opacity
  from the WVM as a proxy for atmospheric emission (arbitrary
  scale). Bolometer baseline drifts on timescales $\gsim$100\,s are
  clearly correlated with the atmosphere in this case.}
\label{fig:skynoise}
\end{figure}

The data shown in Figs.~\ref{fig:bolos_mix} and \ref{fig:pspec} were
taken during fairly stable weather conditions (although with different
450\,\micron\ line-of-sight opacities $\sim$0.9 and 6.5 in the S2SRO
and post-upgrade data sets, respectively), and are dominated at
low frequencies ($\lsim$2\,Hz) by scan-synchronous noise, rather than
underlying variations in the atmosphere, or ``sky noise''. However,
only $\sim$100\,s of data were analysed. In Fig.~\ref{fig:skynoise},
an order-of-magnitude longer (fully-commissioned) data set is plotted
from observation 18 on 2011-03-24, during highly variable weather
conditions (450\,\micron\ line-of-sight opacities ranging from 3.6 to
7.7, scaled from the WVM data which are shown as a black line). Unlike
the earlier plots, it is clear that the bolometer baseline drifts on
timescales $\gsim$100\,s are tightly correlated with the atmospheric
opacity, which serves as a proxy for atmospheric emission. The weaker
correlation between the WVM and the 450\,\micron\ bolometers is
probably due to the atmosphere being more transmissive at 850\micron,
therefore the emission detected at 450\,\micron\ comes from water
vapour that is closer to the telescope.

@@

The accurate measurement of positions and brightnesses of known point
sources are necessary in real-time to establish telescope pointing
offsets and focus. They are also necessary to measure the FCF
(absolute calibration), and hence noise performance of the instrument
in astronomically-useful units.  In this example we reduce a
450\,\micron\ map of Uranus (observation 26 on 2011-10-17), which is a
nearly point-like source for \scuba\ that is commonly used as a
primary flux calibrator. The CV daisy pattern was used, with a scan
speed of 155 arcsec\,sec$^{-1}$. We perform several different
reductions of the data to illustrate the purpose of various model
components and the convergence properties of the solution
(Fig.~\ref{fig:pointmaps}). In all cases the maps are produced on a
grid of azimuth (horizontal) and elevation (vertical) offsets from the
position of Uranus (the origin), using 2\,arcsec pixels.

@@

\scuba\ surveys designed to detect extremely faint point-sources
(e.g., high-redshift star-forming galaxies, and features in debris
disks) are ideally limited by the white-noise performance of the
instrument. The approach described here for maximising the \snr\ of
point-sources involves three major steps: (i) generating a map that
removes most low-frequency noise sources with approximately linear
response, without prior knowledge of the location of sources; (ii)
apply a Fourier-space ``whitening filter'' to suppress residual
large-scale noise; and (iii) detecting point sources using a ``matched
filter''. Note that variations on this general procedure have been
used extensively in the submm cosmology community using previous
instruments
\citep[e.g.,][]{scott2002,borys2003,laurent2005,coppin2006,scott2008,perera2008,devlin2009}.
In this section we reduce scans of the Lockman Hole taken during S2SRO
as a pilot project for the Cosmology Legacy Survey. It consists of
$\sim8.5$\,hours of data taken in 36 separate scans (average
15\,min. each) spread over February and March 2010. Each scan is a
360\,arcsec rotating PONG pattern, with a scan speed of
240\,arcsec\,s$^{-1}$, covering an area of about 50\,arcmin$^2$. The
full list of dates and observation numbers includes: 2010-02-18, 63, 64,
65, 70, 71, 72, 90, 91, 92, 97, 98, 99; 2010-02-20, 111, 112, 113, 118,
128, 129, 130; 2010-03-03, 61, 62, 64, 69, 70, 72, 73, 74, 75; 2010-03-09,
87, 88, 90, 91; and 2010-03-11, 59, 64, 65, 72. These observations
represent about 80 per cent of the total data taken as part of the
project; the remaining observations were dropped due to problems
keeping the arrays properly tuned during S2SRO, and were easily
identified by their highly variable and erratic bolometer time-series
(which resulted in maps full of large streaks).

@@

In this final example, we analyse a map of M17 which contains bright,
extended emission. The data are from observation 11 on 2011-05-31 using
the 850\,\micron\ array. It is a rotating PONG scan covering a
diameter of 0.375\,deg, with a scan speed of 300\,arcsec\,s$^{-1}$,
and a transverse spacing of 180\,arcsec, taking 37.5\,min to complete.

@@

The James Clerk Maxwell Telescope is operated by the Joint Astronomy
Centre on behalf of the Science and Technology Facilities Council of
the United Kingdom, the Netherlands Organisation for Scientific
Research, and the National Research Council of Canada. Additional
funds for the construction of \scuba\ were provided by the Canada
Foundation for Innovation.  This research was supported in part by the
Natural Sciences and Engineering Research Council of Canada.  EC
thanks CANARIE/CANFAR for additional funding.  The authors thank the
members of the SCUBA-2 commissioning team for testing the map-maker
and reporting anomalies; in particular Antonio Chrysostomou and
Jessica Dempsey.  We also thank Mark Halpern, Matthew Hasselfield, and
Gaelen Marsden for many useful discussions; observers who provided
helpful feedback, especially David Nutter and Todd McKenzie; and
Mandana Amiri and Dan Bintley for consultations regarding the SCUBA-2
Multi-Channel Electronics.  We acknowledge the contributions of Dennis
Kelly, Alex van Engelen and Jennifer Balfour for early investigations
related to SMURF; and Mark Thompson, Craig Walther and S\'{e}verin
Gaudet for being on the Critical Design Review panel. We thank Per
Friberg and Gary Davis for their helpful comments on the
manuscript. Finally, we thank the anonymous referee for a very helpful
review that helped to clarify several key areas of the paper.

>>>>

\caption{A comparison between single bolometer time-series in each of
  the 450 and 850\,\micron\ bands with the mixing chamber temperature,
  the line-of-sight 450\,\micron\ opacity derived from the JCMT water
  vapour monitor, and azimuth/elevation pointing offsets, before and
  after upgrading the instrument. The grey signals over-plotted in the
  top two panels show the residual time-series after removing the
  common-mode signals.  (a) Data taken during the S2SRO period,
  observation 29 on 2010 March 13. There is a strong correlation
  between the bolometers and the roughly $\sim$25\,s oscillation in
  the fridge, but only a minor correlation with the opacity and
  telescope motion. The nearly flat common-mode subtracted signals
  show: (i) that most of the low-frequency signal is common to all of
  the bolometers; and (ii) the non-correlated, and predominantly white
  noise at 450\,\micron\ is significantly larger than at
  850\,\micron. (b) Data taken with the fully-commissioned instrument,
  observation 38 on 2011 November 12. Unlike the S2SRO data, there is
  no strong signal produced by variations in the fridge
  temperature. The 450\,\micron\ data have minimal correlated
  low-frequency noise (as evidenced by the lack of a difference
  between the raw and common-mode subtracted signals), although the
  850\,\micron\ data show a common signal correlated with the opacity
  and telescope motion (note that this scan has a significantly larger
  amplitude than the S2SRO data set), mostly likely caused by a
  combination of changing airmass (anti-correlated with opacity), and
  magnetic field pickup (see Section~\ref{sec:magpickup} and
  Fig.~\ref{fig:magpickup}).}
\label{fig:bolos_mix}
\end{figure*}

@@

In the S2SRO data (Fig.~\ref{fig:bolos_mix}a), observation 29 on 2010
March 13, both bolometers share significant long-timescale structure
($\gsim10$\,s) that appears to be related to variations in the fridge
base temperature, although the similarity is clearly greater at
850\,\micron. Note that the total power in the fluctuations at
450\,\micron\ are comparable to those at 850\,\micron\ as one might
expect if there is a comparable varying thermal load from the fridge
at each wavelength that dominates. We also note that there is no
obvious strong correlation between the low-frequency signal structure,
at either wavelength, with the telescope motion. However, there is a
suggestion that the shorter-timescale behaviour of the opacity is
anti-correlated with the elevation, as expected.

@@

Data from the fully commissioned instrument, observation 38 on 2011
November 12, are shown in Fig.~\ref{fig:bolos_mix}b. Having solved the
fridge oscillation problem, these data no longer exhibit a correlation
with the mixing chamber (however, note that variations are still seen
in the mixing chamber signal; such variations do not necessarily
reflect changes in the focal plane temperature). There is minimal
correlated low-frequency noise in these 450\,\micron\ data, resulting
in little difference between the raw and common-mode subtracted
data. The 850\,\micron\ channel, however, exhibits a more significant
signal that is obviously correlated with the telescope motion. It is
also correlated amongst many of the detectors, and common-mode removal
corrects it to a large extent. Part of the reason that this is seen
here, and not in the S2SRO data set shown, is that the amplitude of
the scan pattern is larger. This ``scan-synchronous'' noise is
attributed to a combination of magnetic field pickup, as described in
Section~\ref{sec:magpickup}, and sky brightness variations due to
changes in elevation (not to be confused with underlying changes in
the atmosphere; a homogeneous atmosphere will appear brighter with
increasing airmass). Also similar to the S2SRO data, an
anti-correlation between the opacity and elevation is apparent.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{magpickup}
\caption{Evidence for significant magnetic field pickup for
  observation 16 on 2010 February 28 at 450\,\micron\ (s4a subarray).
  The top panel shows two un-flatfielded (but mean-subtracted and step
  corrected) bolometer time-series from the same column, with a 200
  sample boxcar smooth (approximately 1\,s), illustrating that they
  are dominated by a similar signal with opposite signs. The second
  panel shows the dark squid signal for the column, also
  mean-subtracted and with the same boxcar smooth. The bottom panel
  shows the azimuthal and elevation offsets from the map centre (mean
  azimuth and elevation 171.9$^\circ$ and 68.0$^\circ$,
  respectively). Only the azimuthal signal is obviously correlated
  with the dark squids and bolometer signals, which suggests a
  magnetic field stationary with respect to the telescope dome as the
  source, since its direction with respect to the cryostat only
  changes with azimuthal motion.}
\label{fig:magpickup}
\end{figure}

@@

Example data from the 450\,\micron\ subarray s4a where pickup appears
to be significant (observation 16 on 2010 February 28) are shown in
Fig.~\ref{fig:magpickup}. The time-series for two bolometers in the
same column (not flatfielded) show that there is a strong signal with
a similar shape, but opposite signs. This behaviour is seen across the
entire array. The dark squid signal for the same columns exhibits a
similar shape and amplitude. Since the dark squid has no thermal
absorber or TES attached to it, this observed signal is not likely to
be optical or thermal in nature (although there can be some cross-talk
with the bolometers). Due to the fact that the sign of the gain in
each stage of SQUID amplification is random (although the combined
gain is constrained to be negative), and since magnetic field pickup
is only seen at the input to the second stage, the pickup can appear
with random signs for bolometers along a column, giving it a distinct
signature from other common signals that always appear with the same
sign.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{skynoise}
\caption{Illustration of the dominance of atmospheric variations on
  long-timescales using 35\,min. of data taken in highly-variable
  weather conditions (line-of-sight 450\,\micron\ opacity ranging from
  3.6 to 7.7) from the fully-commissioned instrument, observation 18
  on 2011 March 24. Red and blue lines show the 850 and 450\,\micron\
  common-mode signals, corresponding to the left- and right- vertical
  axes, respectively. The black line shows the line-of-sight opacity
  from the WVM as a proxy for atmospheric emission (arbitrary
  scale). Bolometer baseline drifts on timescales $\gsim$100\,s are
  clearly correlated with the atmosphere in this case.}
\label{fig:skynoise}
\end{figure}

The data shown in Figs.~\ref{fig:bolos_mix} and \ref{fig:pspec} were
taken during fairly stable weather conditions (although with different
450\,\micron\ line-of-sight opacities $\sim$0.9 and 6.5 in the S2SRO
and post-upgrade data sets, respectively), and are dominated at low
frequencies ($\lsim$2\,Hz) by scan-synchronous noise, rather than
underlying variations in the atmosphere, or ``sky noise''. However,
only $\sim$100\,s of data were analysed. In Fig.~\ref{fig:skynoise},
an order-of-magnitude longer (fully-commissioned) data set is plotted
from observation 18 on 2011 March 24, during highly variable weather
conditions (450\,\micron\ line-of-sight opacities ranging from 3.6 to
7.7, scaled from the WVM data which are shown as a black line). Unlike
the earlier plots, it is clear that the bolometer baseline drifts on
timescales $\gsim$100\,s are tightly correlated with the atmospheric
opacity, which serves as a proxy for atmospheric emission. The weaker
correlation between the WVM and the 450\,\micron\ bolometers is
probably due to the atmosphere being more transmissive at 850\micron,
therefore the emission detected at 450\,\micron\ comes from water
vapour that is closer to the telescope than that observed at the
longer wavelengths.

@@

The accurate measurement of positions and brightnesses of known point
sources are necessary in real-time to establish telescope pointing
offsets and focus. They are also necessary to measure the FCF
(absolute calibration), and hence noise performance of the instrument
in astronomically-useful units.  In this example we reduce a
450\,\micron\ map of Uranus (observation 26 on 2011 October 17), which
is a nearly point-like source for \scuba\ that is commonly used as a
primary flux calibrator. The CV daisy pattern was used, with a scan
speed of 155 arcsec\,sec$^{-1}$. We perform several different
reductions of the data to illustrate the purpose of various model
components and the convergence properties of the solution
(Fig.~\ref{fig:pointmaps}). In all cases the maps are produced on a
grid of azimuth (horizontal) and elevation (vertical) offsets from the
position of Uranus (the origin), using 2\,arcsec pixels.

@@

\scuba\ surveys designed to detect extremely faint point-sources
(e.g., high-redshift star-forming galaxies, and features in debris
disks) are ideally limited by the white-noise performance of the
instrument. The approach described here for maximising the \snr\ of
point-sources involves three major steps: (i) generating a map that
removes most low-frequency noise sources with approximately linear
response, without prior knowledge of the location of sources; (ii)
apply a Fourier-space ``whitening filter'' to suppress residual
large-scale noise; and (iii) detecting point sources using a ``matched
filter''. Note that variations on this general procedure have been
used extensively in the submm cosmology community using previous
instruments
\citep[e.g.,][]{scott2002,borys2003,laurent2005,coppin2006,scott2008,perera2008,devlin2009}.
In this section we reduce scans of the Lockman Hole taken during S2SRO
as a pilot project for the Cosmology Legacy Survey. It consists of
$\sim8.5$\,hours of data taken in 36 separate scans (average
15\,min. each) spread over February and March 2010. Each scan is a
360\,arcsec rotating PONG pattern, with a scan speed of
240\,arcsec\,s$^{-1}$, covering an area of about 50\,arcmin$^2$. The
full list of dates and observation numbers includes: 2010 February 18,
63, 64, 65, 70, 71, 72, 90, 91, 92, 97, 98, 99; 2010 February 20, 111,
112, 113, 118, 128, 129, 130; 2010 March 3, 61, 62, 64, 69, 70, 72,
73, 74, 75; 2010 March 9, 87, 88, 90, 91; and 2010 March 11, 59, 64, 65,
72. These observations represent about 80 per cent of the total data
taken as part of the project; the remaining observations were dropped
due to problems keeping the arrays properly tuned during S2SRO, and
were easily identified by their highly variable and erratic bolometer
time-series (which resulted in maps full of large streaks).

@@

In this final example, we analyse a map of M17 which contains bright,
extended emission. The data are from observation 11 on 2011 May 31
using the 850\,\micron\ array. It is a rotating PONG scan covering a
diameter of 0.375\,deg, with a scan speed of 300\,arcsec\,s$^{-1}$,
and a transverse spacing of 180\,arcsec, taking 37.5\,min to complete.

@@

The James Clerk Maxwell Telescope is operated by the Joint Astronomy
Centre on behalf of the Science and Technology Facilities Council of
the United Kingdom, the Netherlands Organisation for Scientific
Research, and the National Research Council of Canada. Additional
funds for the construction of \scuba\ were provided by the Canada
Foundation for Innovation.  This research was supported in part by the
Natural Sciences and Engineering Research Council of Canada.  EC
thanks CANARIE/CANFAR for additional funding.  The authors thank the
members of the SCUBA-2 commissioning team for testing the map-maker
and reporting anomalies; in particular Antonio Chrysostomou and
Jessica Dempsey.  We also thank Mark Halpern, Matthew Hasselfield, and
Gaelen Marsden for many useful discussions; observers who provided
helpful feedback, especially David Nutter and Todd McKenzie; and
Mandana Amiri and Dan Bintley for consultations regarding the SCUBA-2
Multi-Channel Electronics.  We acknowledge the contributions of Dennis
Kelly, Alex van Engelen and Jennifer Balfour for early investigations
related to SMURF; and Mark Thompson, Craig Walther and S\'{e}verin
Gaudet for being on the Critical Design Review panel. We thank Per
Friberg and Gary Davis for their helpful comments on the
manuscript. Finally, we thank the anonymous referee for their thorough
review that helped to clarify several key areas of the paper.
